---
title: "Activity Monitoring"
author: "Don Thapa"
date: "February 26, 2016"
output: html_document
---
```{r setoptions, echo = FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE)
```

This report looks at some activity monitoring data collected at 5 minute interval  
througout the day for an anonymous individual during the months of October and November 2012.

#### Loading and preprocessing the data
```{r}
setwd('~/Documents/RepData_PeerAssessment1/')
activity <- read.table(unz('activity.zip', 'activity.csv'), 
                       header = TRUE, sep = ",")
library(lubridate)
activity$date <- ymd(activity$date)
str(activity)
```

#### What is the mean total number of steps taken per day
```{r}
library(dplyr)
stepsByDay <- group_by(activity, date) %>% 
    summarise(sum(steps))
names(stepsByDay) <- c("date", "totalSteps")
meanSteps <- mean(stepsByDay$totalSteps, na.rm = TRUE)
medianSteps <- median(stepsByDay$totalSteps, na.rm = TRUE)
library(ggplot2)
ggplot(stepsByDay, aes(totalSteps)) + 
    geom_histogram(col = "blue", fill = "orange", binwidth = 1000) + 
    geom_vline(data = stepsByDay, aes(xintercept = meanSteps), 
               col = "red", size = 1.5) +
    labs(title = "histogram of average steps per day", x = "steps")
summaryActivity <- summary(activity$steps)
```

The mean and median steps taken per day is 
```{r echo=FALSE}
meanSteps
```
and
```{r echo=FALSE}
medianSteps
```
respectively.

#### What is the average daily activity pattern?
```{r}
by_interval <- group_by(activity, interval) %>% 
    summarise(mean(steps, na.rm = TRUE))
names(by_interval) <- c("interval", "meanSteps")
maxMeanSteps <- by_interval[which.max(by_interval$meanSteps), ]
ggplot(by_interval, aes(interval, meanSteps)) + 
    geom_line(col = "blue") + 
    geom_point(alpha = .5, col = "red") +
    geom_text(data = maxMeanSteps, 
              aes(label = paste("highest at", round(interval, 1), ",", 
                                round(meanSteps, 1))), 
              size = 5, col = "purple") + 
    geom_smooth(method = "lm", col = "orange") +
    geom_hline(aes(yintercept = mean(activity$steps, na.rm = TRUE)), 
               col = "lightblue") + 
    geom_point(data = maxMeanSteps, aes(interval, meanSteps), 
               alpha = .4, size = 8, col = "orange") +
    labs(title = "average steps per 5 min interval", 
         y = "mean steps", x = "5 min intervals")
```

The interval with the highest mean steps of **`r maxMeanSteps$meanSteps`** is at **`r maxMeanSteps$interval`**

#### Imputing missing values
```{r}
missingValues <- sum(!complete.cases(activity))
```
The data contains **`r missingValues`** missing values

A simple approach to imputating missing values is adopted where missing steps are  
simply replaced by the mean steps without any grouping calculated below **`r mean(activity$steps, na.rm = TRUE)`**

```{r}
activity$steps[is.na(activity$steps)] = mean(activity$steps, na.rm = TRUE)
stepsByDay_imputed <- group_by(activity, date) %>% 
    summarise(sum(steps))
names(stepsByDay_imputed) <- c("date", "totalSteps")
meanStepsImputed <- mean(stepsByDay_imputed$totalSteps)
medianStepsImputed <- median(stepsByDay_imputed$totalSteps)
ggplot(stepsByDay_imputed, aes(totalSteps)) + 
    geom_histogram(col = "blue", fill = "orange", binwidth = 1000) + 
    geom_vline(data = stepsByDay_imputed, aes(xintercept = meanStepsImputed), 
               col = "red", size = 1.5) +
    labs(title = "average steps per day", x = "steps")
```

The difference in mean steps and median steps of the imputed data and the orignal are now **`r meanStepsImputed - meanSteps`** and **`r medianStepsImputed - medianSteps`** respectively.  
These two statistics essentially remained the same before and after imputation, but it is clear from  
the following boxplot, that imputation has decreased the interquartile range.

```{r}
boxplot(summaryActivity, summary(activity$steps))
```

#### Are there differences in activity patterns between weekdays and weekends?
```{r}
library(chron)
activity$weekend <- is.weekend(activity$date)
activity$day <- "weekday"
activity$day[activity$weekend == T] <- "weekend"
by_day <- group_by(activity, interval, day) %>% 
    summarise(mean(steps, na.rm = TRUE))
names(by_day) <- c("interval", "day", "meanSteps")
maxMeanSteps <- by_day[which.max(by_day$meanSteps), ]
ggplot(by_day, aes(interval, meanSteps)) + 
    geom_line(col = "blue") + 
    geom_point(alpha = .5, col = "red") +
    geom_smooth(method = "lm", col = "orange") +
    facet_grid(. ~ day) +
    labs(title = "average steps per 5 min interval", 
         y = "mean steps", x = "5 min intervals")
```

Lets look at the summary next.
```{r}
weekend <- filter(activity, day == "weekend") %>%
    group_by(date) %>% summarise(sum(steps))
names(weekend) <- c("date", "totalSteps")
weekday <- filter(activity, day == "weekday") %>%
    group_by(date) %>% summarise(sum(steps))
names(weekday) <- c("date", "totalSteps")
summary(weekend$totalSteps)
summary(weekday$totalSteps)
```

So it does seem like there is some distinction between weekend and weekday mean steps - ***`r mean(weekend$totalSteps)`*** for weekend vs ***`r mean(weekday$totalSteps)`*** for weekdays. 

#### Conclusion
Lets validate what we saw above using a two sided t-test next using unequal variance.

```{r results='asis'}
t.test.result <- t.test(weekend$totalSteps, weekday$totalSteps)
```
```{r echo=FALSE}
t.test.result
```

Given that the 95% confidence interval of the differences between the two subsets is **`r t.test.result$conf.int`** (so doesn't include 0) and  
the t-statistic is **`r t.test.result$statistic`** which is much higher than the 95% confidence t-statistic for a **`r t.test.result$parameter`** degrees of freedom of around  
**`r qt(.975, t.test.result$parameter)`** and the associated p-value of the t test is **`r t.test.result$p.value`** (assuming alpha at 5%), we can statistically conclude that the individual walked on  
average more on the weekends than weekdays.
